\thetitle

### 2021-02-10 writing/reviews/pldi-rebuttal-checklist.md

- Addresses a problem that has not been addressed before.
- Clear, concise, and meaningful methodology.
- Clear description of the threats to validity.
- Very well-written paper.

Points against:
- none

# Questions

DONE
- Could you comment a bit more on what makes a mutator "good"? (other than
  experimentally seeing that it is good.) Perhaps, what would be a really bad
  mutator for evaluating blame assignment?

DONE
- I don't understand where the 16,800 number comes from. Is it an exhaustive
  enumeration of all possible mutations? (Or the number of mutants constructed
  in X amount of time?)

DONE
- How large are the programs in the benchmark suite?

@@

# Minor Comments

DONE
- L38: Should there have been a colon? Is the text "This contrast... " and
  challenge quote from [3]? Who is saying what here...

DONE
- L104: I was confused by the word "Next". Maybe instead: We now ..."

DONE
- L268: Grammar issue.

DONE
- L314: Floating period.

DONE
- L441: Add line numbers (or other metric of size).

DONE
- L956: 30,000 hours of compute time is 1,250 days. I assume that the CPU has
  many cores. Could you put that number?

DONE
- L980: Figure 5.4??

DONE
- L991: Figure 6: It took me some time to understand the figure. I wonder if
  there is a better way to represent the same data?

@@

TODO: This seems to imply that we are proposing or assuming the way that real "rational programmers" would behave. Should we try to answer or clarify about that?

Comments for author
-------------------
This paper tackles a long-standing difficult problem, which is the
problem of evaluating blame. Many existing gradual typing papers
provide blame systems, but because the papers do not evaluate blame
systems, it is not obvious whether blame itself is indeed helpful for
debugging. Among many theoretical gradual type systems adopting blame
assignments, no paper discussed the evaluation of blame systems. It is an
important problem because, without any evaluation, theoretical aspects
like gradual guarantee and the blame theorem may not be applicable to
practical systems combining static and dynamic typing.

This paper shows that blame assignments do help programmers, and it
may serve as a first step to close the gap between the theory and
the practice of gradual typing.

DONE: this is directly addressed in minor points, and vaguely by the high lvl response
The experiment results show that there are debugging scenarios where
Transient is more useful than Natural. Does it happen simply because
of luck? Or, does it imply that Transient is more suitable than
Natural for a particular kind of bugs for some reason? Also, can there
be a new blame strategy that outperforms both Transient and Natural
all the time?

DONE
Theoretical gradual type systems usually allow programmers to control
the precision of type annotations in a fine-grained way with the Dyn
type rather than distinguishing typed and untyped modules. Would the
same evaluation method be able to directly appliable to blame in such systems?

DONE
Please explain what impedance mismatches are in the paper. Is it
simply type mismatches?

@@

# Strengths

* The paper is well-organized.
* It proposes a novel solution to a long-standing difficult problem.
* Threats to validity are discussed in detail.

# Minor comments

DONE
The paper states that the effort distribution of the random mode
follows a normal distribution, as expected. Please explain why a
normal distribution is a reasonable expectation. When there are $n$
untyped components, the random mode randomly selects one of them, and
only one of them is a buggy component. Therefore, in the first
iteration, the random mode reaches the end of the trail with the
probability of $1/n$. In the second iteration, there are $(n-1)$
untyped components and the probability of a failure at the first
iteration is $(n-1)/n$, so the probability of effort $2$ is $1/n$
again. In this way, it seems to be that the probability of effort $i$
is $1/n$ where $1<=i<=n$ and $n$ is the number of initially untyped
components. Each trail has a different number of untyped components,
so the resulting distribution would be the average of such
distributions. Based on this reasoning, it is unclear why the
distribution is a normal distribution. This point does not affect the
contribution of the paper, but it is not clear from the paper.

DONE
* page 9 line 980: Figure 5.4 -> Figure 6
* page 10 Figure 7: Null -> Random
* page 11 line 1103: ( 770 -> (770

@@

To evaluate this approach, the paper also presents empirical results. Starting with a small corpus of programs, the authors use mutations to systematically (but synthetically) introduce type errors -- e.g., one mutation replaces a constants with another of a different type, and another mutation swaps method identifiers. They then conduct experiments to quantify the number of steps required by a rational programmer for several blame assignment schemes: Natural (used in Typed Racket, which tracks at most one location for blame), Transient (used in Reticulated Python, which tracks a set of locations), and Erasure (which does not actually assign blame, and is widely used in industry).
TODO: do we need to clarify that effort is a secondary aspect of the evaluation, and more important is whether blame leads to the bug at all? This is not explicitly clarified anywhere in the rebuttal.

Unsurprisingly, the evaluation shows that blame is effective, when used by a rational programmer. The results are inconclusive, but some trends are clear. Most significant, all of the schemes that assign blame are more useful than Erasure. And Natural blame more often more useful than Transient blame.

Comments for author
-------------------
This is a neat paper that develops an approach for evaluating gradual typing and blame assignment. Thanks for submitting it to PLDI '21! The results are presented in an extremely clear way that draws out the key insights without compromising on technical detail. The notion of a rational programmer -- loosely modeled on rational agents from economic theory -- is a neat conceptual device. And the empirical approach does seem like the right way to address the question of the usefulness of blame.

But while there is a lot to like about the paper, it also has some shortcomings. The suite of programs used for the evaluation is fairly small (c.f., the SIGPLAN checklist).
DONE ^
The faults are introduced using synthetic mutations that may or may not correspond to the kinds of errors that arise in practice. This can be seen from the results in Figure 7 -- the lengths of the "trails" is quite small.
DONE ^
While the study is mostly well done, it only considers two blame assignment strategies.
DONE ^
For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame. But this is not done.
DONE ^
And unfortunately the results are inconclusive -- Section 8 says "our results call for a deeper understanding of the two models for blame".
DONE ^
Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
DONE ^
Most if not all of these shortcomings are discussed in the paper, which is much appreciated. But they do risk undermine the value of the contributions to some degree.

DONE
To explain one comment above: I was surprised not to see a comparison against an "omniscient programmer" to establish a baseline (c.f., the SIGPLAN checklist). That is, model a blame assignment scheme in which blame could be placed on any component. And always pick the component that minimizes the length of the trail. Of course, this would be infeasible to implement in practice, but it would place an upper bound on the utility of blame.

DONE
Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices. Section 5.2 just says "our answer" is that there are at least "two reasonable options." Why are other choices not reasonable?

@@

The current paper extends the POPL 2020 methodology from contracts in
Racket to gradual (or migratory) typing in Typed Racket, in several
steps:

First, to complement the "Natural" higher-order contract system of
(Typed) Racket, this work implements a "Transient" blame assignment
strategy, which unlike Natural inserts only shallow run-time checks.
(To date, Transient has appeared only in the separate design and
implementation of Reticulated Python, which prevents a head-to-head
comparison as desired in this work.)

Second, compared to POPL 2020, new mutators are defined to inject the
kinds of type-mismatch bugs that may be caught by the gradual type
system and corresponding run-time checks and blame assignment.

Third, the paper runs a large experiment on 10 benchmark programs,
inserting mutations and analyzing whether the blame trails produced by
each debugging mode --- following (a) the component blamed by Natural,
(b/c) the first or last component, respectively, blamed by Transient,
or (d) an exception raised by the underyling language --- conclude
with the actual bug. Results show that Natural, Transient-First, and
Transient-Last all produce short successful blame trails (typically of
length one or two) on most debugging tasks, and that they are all
similarly effective with no clear winner for all tasks.

@@

Comments for author
-------------------
This work advances a compelling approach to evaluating a usability
question through a systematic exploration of synthetic bugs and fixes.
The large-scale experiment is impressive, and I find both of the main
results surprising:

1. That Natural is not much more effective than Transient. I thought
the shallow checks of Transient would certainly be less effective than
the full checks of Natural. It will be interesting to see if and how
this holds up in subsequent usability studies, both with synthetic and
real users and tasks.

DONE
2. That Transient --- as implemented in Typed Racket --- is much
slower than Natural. This also goes against intuition, as well as
previous results reported about Reticulated Python (but apparently
there were flaws in those experiments, as noted in this paper). Given
that the experiments are developed and run in the context of the
large, real-world Racket ecosystem, I wonder whether about the chance
of incidental implementation choices or bugs affecting this result.

Overall, I think this is a useful reference for further usability work
on contracts and gradual type systems, and am generally in favor of
acceptance.

@@

At the same time, however, I believe the paper overstates the novelty
and challenges to "bring [the POPL 2020 approach] to the world of
gradual typing" (L1252). Comments about each of the three main
components of the methodology in turn below.

DONE
*Natural and Transient Under the Same Roof (Section 3)*

@@

Review #66E
===========================================================================

Overall merit
-------------
C. I would not accept this paper but will not argue strongly against
   accepting it.

Reviewer expertise
------------------
Z. **Some familiarity** = "I have a passing knowledge of the topic(s) but
   do not follow the relevant literature"

Paper summary
-------------
DONE: clarifying that we propose a methodology, and what the key experimental results are
This paper seeks to establish (or refute) the benefit of blame tracking in gradually typed programming environments. It takes several Typed Racket benchmarks and considers multiple approaches to blame tracking. The central idea is to use a fault injection method to create variant programs along two dimensions: type errors inserted according to some crafted heuristics, and static typing selectively removed on a component-by-component granularity. This creates a large number of variant programs which may produce run-time type errors with blame; this is then iterated to simulate a "rational programmer", i.e. such that blame is used to select where more static checking will be added next; at any given iteration this may or may not catch the injected error, and the number of iterations is counted as a proxy for effort. A further contribution is implementing the different forms of blame-tracking in the same framework so that they can be compared. The finding is that blame is found to offer some positive advantage in effort (over simply being guided by the stack trace of a run-time exception) in 9--13% of the initial sample of variants, with some minor differences between the forms of blame tracking tried.

@@

Comments for author
-------------------
I commend the authors for asking the question. Their method is clearly the product of much careful thought. I understand the difficulty in doing experiments about errors that tend not to hit 'real' source code repositories, and I appreciate the effort taken to re-create the different approaches to blame in a comparable setting.

DONE
However, I don't find the results to be as conclusive as the authors state. Aside from the Erasure case, whose relevance I found unclear (see below), conclusions rest on some small-valued "more useful than" percentages. These are pretty hard to interpret, because the "usefulness" metrics necessarily build in a lot of simplifying assumptions. E.g. in 5.4, the "percentage of scenarios where ... is more effective" doesn't account for *extent* of difference, and the trail-length "programmer effort" metric in 5.5 is also unlikely to be a great proxy. These limitations are understandable, method-wise, but overall the results are fairly null... they let us continue believing "what we'd expect" but not with any great sense of added confidence. That doesn't entirely diminish the work, of course, but I'd say the current write-up overplays its hand a little. I appreciate the discussion of threats to validity.

@@

The presentation of the results is very much around aggregates and summaries. Indeed the whole method is about having run a huge compute job over a large number of variants. Perhaps it's paranoia but I'd be interested to see some specific examples walked through by hand (in the paper) and some smaller sample manually classified (as results). That would add an extra sanity check that the metrics do correspond to some meaningful reality.

DONE
About Erasure: if I'm reading this correctly, the key thing here is that in Typed Racket, the set of static types is more expressive than the set of dynamic types. For example, there is a static notion of non-negative integer that is distinct from plain integer. This sort of design isn't universal -- in some languages/systems static types closely mirror the classification of objects in the language's dynamic semantics. Writing these stricter contracts into a program brings its own benefits, separate from blame or indeed from static checking. It feels like the paper doesn't take enough care to distinguish the two effects: the effect of systematically applying more refined contracts over program values, and the effect of gradually enabling static checking of those contracts (iteratively, guided by blame). The authors do mention this around line 204, and the comparisons in Figure 6 between "_ exceptions" and "Erasure" seem to be measuring this -- the gain from checking these extra annotations, with the more precise and/or more timely checks that they imply, relative to the erased case where only the language baseline contracts are checked. Indeed that's the point of 'exception' experiments. Since the biggest effect sizes on display are these ones -- between Erasure and anything else -- this seems at best distracting. I'd be glad to hear from the authors if I'm misunderstanding anything here. In the detailed comments below, I have noted some places in the text where it would be useful to remind the reader that this design property of Typed Racket is at play.

DONE
It's interesting to note that there seems to be no convincing analogous experimental or modelling-based justification for plain old static typing, i.e. showing that it somehow presents a net gain to the programmer. Rather, this has simply been posited/assumed by a very long line of work. I am not defending that state of affairs, but it points to the difficulty of showing conclusively that something truly helps programmers. I can see value in this sort of simulation-style approach, and it is no less convincing than user studies. So I'd be interested to hear from the author(s) if they have any more arguments that (1) I've underestimated the results' conclusiveness, or (2) this is a novel/interesting family of methods that might be pursued more widely, or (3) that deeper experiments building on these ideas might yield more compelling insights.

@@

Detailed comments:

TODO
In the title, "evaluating blame" reads oddly. It is ambiguous, because "evaluate" sometimes means to compute a result. A fuller phrase, like "evaluating the usefulness of blame tracking...", would probably be worth the words.

TODO
The abstract doesn't say much about the work. It would be better written for experts to quickly gather an overview what the paper contributes.

TODO
Conversely, the main body of the paper skimps a bit on background. It never explicitly covers what "blame" means and how it works in practice. Similarly, it repeatedly talks about "impedance mismatches" without defining them. Perhaps this phrase is now standard in the gradual typing literature, which I haven't kept up with (hence my Z expertise). But I did read the Wadler/Findler paper carefully at the time. It did not talk about impedance mismatches. In any case, it is a fuzzy metaphor... please say exactly what it means here.

DONE
line 34: "then their compilers remove types and rely on the built-in safety checks of the underlying language to catch any problems". This reads oddly in context. Didn't they just do a bunch of static checking? So they are *not* relying just on dynamic checks to catch problems? Maybe there is something more accurate to say here... e.g. no dynamic check is removed, or something like that.

DONE
line 38: "explicit statement and challenge" -- what is it?

DONE
line 49: I agree they got the word wrong, but you should explain this

DONE
line 60: around this paragraph the writing started to grate. There's no need to generalise about what people do or think, and the "As a matter of fact... simply..." style is somewhere between laboured and patronising. It's better to more plainly state the gap in the literature that you're addressing. Throughout the paper, much space could be saved by writing in a more direct style.

DONE
line 69: be explicit that Lazarek et al were (as I later gathered) doing something about blame to do with higher-order contracts but not gradual types

DONE
line 70: what does it mean to "follow" the slogan? Was unclear to me.

DONE
line 90: at first I wondered: what is a case? Maybe say "program variant" to foreshadow the idea of generating mutants etc?

DONE
line 91-ish: "Transient, "Natural", "Erasure" -- be explicit that these are names that *you* are introducing

DONE
line 94: "forego" => "forgo"

DONE
line 119: "homo economicus" needs glossing or a reference

DONE
line 125: was wondering whether "impedance mismatch" just means "feasible run-time type error".

DONE
line 148: last sentence appears to contradict the preceding paragraph, and is left hanging oddly. Instead, make it the start of a new paragraph.

DONE
line 160: "disparity" is a strange word here. "diversity"?

DONE
line 170: "Otherwise, results from..." -- the point was already clear

DONE
line 174: ... instead of just saying what you built on, first say what you did! It's really not clear at this point.

DONE
line 187: what is a "component"? Should be easy to define.

TODO
line 199: "blame set as another form of a stack trace" -- yes. I was hoping to see a clearer example of debugging with and without blame, i.e. something making explicit the similarities and differences between having blame info and having only a stack trace at the error site.

DONE
line 211: "languages exceptions" typo

DONE
line 268: this explanation of the workings of proxies seemed overwrought

DONE
line 277: what is "responsibility" of a "party", exactly? Are "party", "module" and "component" all the same things?

DONE
line 283: "rewrites typed modules to inline checks" -- so a module gets turned into a check? Clearly not, but that's how it reads

DONE
line 293: I was wondering what constitutes a boundary crossing. Clearly, passing by function call or return crosses from the caller's module to the callee's. What about values exchanged through reads/writes to shared state?

DONE
line 299: "crosses" => "crossings" (probably)

DONE
line 313: it threw me that a program might not fail but "produce a wrong result". If it could be caught by the gradual type system, why can't it be caught at run time? My best get at explaining this is by what I wrote above, i.e. it's a consequence of Typed Racket's more refined static notion of type. In certain other systems this wouldn't be possible, because the static checker would only catch (albeit earlier) errors that would be caught at run time, so there would be no basis to call the result "wrong".

DONE
lien 352: "three interpretations" -- what are they? I don't see them in the figure.

DONE
line 380: from this I inferred that "migratory typing" means "gradual typing applied at modulewise granularity". Assuming that's correct, it's worth saying directly.

DONE
line 421: "type mistake" -- does this mean "feasible run-time type error"?

DONE
line 424: "fully typed correct programs" -- presumably your method could also work with not-yet-fully-typed correct programs, just not ranging over the entire lattice in those cases. I was wondering whether that might give different/interesting results.

DONE
line 433: "without loss of diversity" -- this is glib. Clearly diversity is lost; just claim that what remains is still diverse enough.

DONE
line 497: "truthiness" needs explaining

DONE
line 517: the difference between #1 and #2 here again relies on the surprising property (to the unfamiliar) that Typed Racket has a notion of "type-level mistake" that doesn't surface under erasure (as an exception or whatever) but also is not a false positive (i.e. you're not talking about conservativeness of static checking). Line 538's "unavoidable" claim is probably also true only in such a context.

DONE
line 519: "at least three" -- better to claim this is only two here, then explain later that the driver doesn't count. Mentioning "three" up-front just raises an unnecessary question in the reader's mind.

DONE
line 634: earlier I had inferred that the distinction between "migratory" and "gradual" was more than just preferring one word; this seems to contradict that

DONE
line 640: what makes it "concise"?

DONE
line 673: "programmers runs" typo

DONE
line 727: "dubbed" -- reads oddly. Maybe italicise the "location", but it seems overkill. The phrase is pretty self-suggesting as it is.

DONE
line 756: "checked the value's type.." -- and the check passed!?

DONE
line 752: "added to the blame set first" -- should it be a blame list, then?

DONE
line 936: no need for hyphen after the adverb

DONE
line 966: "more useful C" -- missing "than"

### 2021-02-10 writing/reviews/pldi-rebuttal-checklist.md

DONE
line 982: don't think these percentages deserve 3 significant figures

DONE
line 1304: "Problem is" -- missing "The". Also, no need to italicise the next sentence... it is really not that deep or surprising.

### 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md

2) Part of  the additional effectiveness of the former is due to runtime
checks (that Erasure foregoes) but another part is due to blame;

3) The effectiveness of the Natural and Transient rational programmers is
comparable.

The first two results confirm what the vast literature on gradual typing
claims without systematic evidence: blame is useful. However, as review
`D` astutely points out, the third conclusion is surprising in two
different ways:

(i) An important motivation for Transient is performance but our
experiment provides evidence that performance issues, which stem from the
design of Transient, negatively impact the effectiveness of blame.

(ii) Theory claims that Natural blame is more accurate than Transient
blame but our experiment provides evidence for the opposite;
ll: TODO is "the opposite" the right phrasing? It seems more accurate to say evidence that doesn't support the prediction.

@@

Challenges for our work compared to Lazarek's POPL paper
===========================================================================

Our work builds on the idea of Lazarek et al.'s paper that it is possible
to evaluate the effectiveness of blame by simulating a debugging process.
Both their work and our work is inspired by Greenman et al's work that
uses a lattice of configurations to systematically explore variants of the
same program with specifications of different precision (for performance
reasons in Greenman et al).  Both the ideas of ``following`` blame while
debugging and of gradually adding specifications to tweak the evaluation
of the program originate from the practice of contracts (see the story in
the intro of Lazarek's paper and similar anecdotes in Meyer's writings).

That said,

1) In the context of Lazarek et al.'s work interesting bugs  are those
detectable by behavioral contracts. Hence, they can use off-the-shelf
mutators. In fact, their contracts do not detect the overwhelming majority
of bugs except at the highest precision, where the contracts describe the full
functional correctness of components.

@@

Implementing Transient Typed Racket is not mere engineering.  Neither
Reticulated nor the Transient model provide answers about what shallow
checks correspond to Typed Racket's sophisticated types (e.g.
polymorphism, row types for mixins, occurrence typing) and what
information is necessary to record in the blame map. Moreover in the
process of the implementation we discovered issues with the published
Transient model and Reticulated that we had to fix. We do not think that
this paper is the appropriate place to describe the implementation in
detail.

2) Custom mutators (section 4).

The identification of good mutators for gradual typing is not
straightforward. A bad mutator results in a type-level bug that is trivial
to locate. For instance, a bug that leads to a type-level error during the
evaluation of the component that contains it is trivial to locate. In
contrast, a bug that is non-trivial to locate is one that causes the buggy
component to evaluate without an error but produce instead a value of the
wrong type. Furthermore an interesting bug is one where the value with the
wrong type manages also to cross undetected to at least another component
before a type-level check flags it. Good mutators should be able to
produce these interesting bugs in our benchmarks. Finally, they should
produce interesting bugs that are detectable by all three systems so that
we can try all systems on the same debugging scenarios.  In general, there
is no methodology for constructing a new set of mutators that produce
interesting bugs for any definition of interesting. So narrowing down the
ones in the paper required rounds of trials to end up with a set of
mutators that leads to a sufficiently large set of interesting and diverse
bugs.

ll: TODO phrasing of "original set we tried" seems off
The sixteen mutators we include in the paper are not the original set we
tried and others that seemed like good candidates proved to be
ineffective. For instance replacing car with cdr should lead to type-level
bugs however none of them are interesting --- they blow up immediately.
Similarly, swapping mutable vectors and hashes with immutable ones is an
interesting mutation in the traditional sense and should lead to a type
error in Typed Racket since its type system distinguishes the two, however
it turns out such mutations  are not interesting in our setting.

@@

It is an exhaustive enumeration. We will clarify the prose.

> - How large are the programs in the benchmark suite?

Here are the LOC counts and number of components for each benchmark. We
will add the information to figure 2.

  benchmark    | LOC  | components
  -------------|------|-----------
  acquire      | 1941 | 9
  gregor       | 2336 | 13
  kcfa         | 328  | 7
  quadT        | 7396 | 14
  quadU        | 7282 | 14
  snake        | 182  | 8
  suffixtree   | 1500 | 6
  synth        | 871  | 10
  take5        | 465  | 8

@@

> - L991: Figure 6: It took me some time to understand the figure. I
>   wonder if there is a better way to represent the same data?

We have reached the same conclusion and we are considering alternatives
that will make it easier to see how two modes compare to each other, such
as a series of two-sided bar charts.

Review #66B
===========================================================================
> The experiment results show that there are debugging scenarios where
> Transient is more useful than Natural. Does it happen simply because of
> luck? Or, does it imply that Transient is more suitable than Natural for
> a particular kind of bugs for some reason? Also, can there be a new
> blame strategy that outperforms both Transient and Natural all the time?

We were also surprised to discover that Transient outperforms Natural in
some scenarios. After all, theory says otherwise. We haven't identified
any patterns for these scenarios. We conjecture that they point to
limitations of the theory but we can't exclude that some of them aren't
because of some unknown Typed Racket bug. Please see the first part of the
discussion at the beginning of our response for more details and for our
take on your last question.

@@

Review #66C
===========================================================================
> But while there is a lot to like about the paper, it also has some
> shortcomings. The suite of programs used for the evaluation is fairly
> small (c.f., the SIGPLAN checklist).

Could you clarify what you mean by fairly small with respect to the
SIGPLAN checklist? We are the first to conduct an empirical study for
blame and gradual typing, so we repurposed the standard performance
benchmark suite for Typed Racket -- in particular, its largest and most
complex benchmarks. Our inspection of the benchmarks indicates that they
are representative set of Racket programs in terms of features and
structure.

> The faults are introduced using synthetic mutations that may or may not
> correspond to the kinds of errors that arise in practice. This can be
> seen from the results in Figure 7 -- the lengths of the "trails" is
> quite small.

There is no existing catalog of bugs in gradually typed programs.
Furthermore, most type-level bugs do not survive  decent testing and thus
do not make it to code repos. There are a few anecdotes from the
literature but they are mostly artificial. Hence, since we need a large
number of buggy scenarios for a meaningful study, a synthetic approach
is the only way forward. In addition, mutation allows us to tune the set
of scenarios to get a diverse set of interesting bugs.  We discuss in
section 8 how our approach to bug injection is a threat to validity.
ll: TODO the review already acknowledges that we discuss this point in the threats; do we need to repeat that?

@@

> While the study is mostly well done, it only considers two blame
> assignment strategies.

We would like to point out that the experiment considers nine different
modes of the rational programmer spread across three different gradual
typing systems.

ll: TODO are we calling the random modes for each of natural, transient, and erasure distinct? I'm not sure that makes sense, in which case it's seven modes.

@@

> Most if not all of these shortcomings are discussed in the paper, which
> is much appreciated. But they do risk undermine the value of the
> contributions to some degree.

We would like to point out that the results of the comparison between the
three systems are only one part of the contribution of the paper. The
other part is the method itself.

> To explain one comment above: I was surprised not to see a comparison
> against an "omniscient programmer" to establish a baseline (c.f., the
> SIGPLAN checklist). That is, model a blame assignment scheme in which
> blame could be placed on any component. And always pick the component
> that minimizes the length of the trail. Of course, this would be
> infeasible to implement in practice, but it would place an upper bound
> on the utility of blame.

Our experiment does have a baseline. The random mode of the rational
programmer serves as the "control" of our experiment. Furthermore the
exception modes each serve as further "controls" to distinguish the effect
of checks from that of blame within each semantics. Please also see our
discussion of modes in the second theme at the beginning of our reply.

@@

> Similarly, I wondered why the modes of the transient programmer only
> consider picking the first/last blame, and not other choices. Section
> 5.2 just says "our answer" is that there are at least "two reasonable
> options." Why are other choices not reasonable?

The Transient semantics provide no interpretation for the blame set. We
picked the first and last because they match the intuitive interpretation
of the blame set as a list. We didn't mean to imply that other options are
unreasonable. We just needed to make a reasonable choice to keep the
experiment feasible.

@@

Review #66D
===========================================================================

> 2. That Transient --- as implemented in Typed Racket --- is much
> slower than Natural. ...
> Given that the experiments are developed and run in the context of the
> large, real-world Racket ecosystem, I wonder whether about the chance of
> incidental implementation choices or bugs affecting this result.

A side note here: Transient Typed Racket is faster than Natural Typed
Racket when we turn off blame. Exactly as a sanity check that this is not
an artifact of our implementation we have coded up benchmarks in Python
and we confirmed that in Reticulated the performance of Transient with
blame is even worse than what we have observed in Typed Racket.

@@

> However, I don't find the results to be as conclusive as the authors
> state. Aside from the Erasure case, whose relevance I found unclear (see
> below), conclusions rest on some small-valued "more useful than"
> percentages. These are pretty hard to interpret, because the
> "usefulness" metrics necessarily build in a lot of simplifying
> assumptions. E.g. in 5.4, the "percentage of scenarios where ... is more
> effective" doesn't account for *extent* of difference, and the
> trail-length "programmer effort" metric in 5.5 is also unlikely to be a
> great proxy. These limitations are understandable, method-wise, but
> overall the results are fairly null... they let us continue believing
> "what we'd expect" but not with any great sense of added confidence.
> That doesn't entirely diminish the work, of course, but I'd say the
> current write-up overplays its hand a little. I appreciate the
> discussion of threats to validity.

We absolutely disagree with you. Please see our discussion of the
conclusions form our experiment at the beginning of our reply.

chrdimo: Should we be more assertive here about the metrics? I think this
is a hard to defend position because of the non-constructive manner of the
criticism --- it is plain dismissive without explanation.

ll: TODO "absolutely disagree with *you*" seems too personal and confrontational. I'm not sure, but just referring to the beginning should speak for itself?

@@

> About Erasure: if I'm reading this correctly, the key thing here is that
> in Typed Racket, the set of static types is more expressive than the set
> of dynamic types. For example, there is a static notion of non-negative
> integer that is distinct from plain integer. This sort of design isn't
> universal -- in some languages/systems static types closely mirror the
> classification of objects in the language's dynamic semantics. Writing
> these stricter contracts into a program brings its own benefits,
> separate from blame or indeed from static checking. It feels like the
> paper doesn't take enough care to distinguish the two effects: the
> effect of systematically applying more refined contracts over program
> values, and the effect of gradually enabling static checking of those
> contracts (iteratively, guided by blame). The authors do mention this
> around line 204, and the comparisons in Figure 6 between "_ exceptions"
> and "Erasure" seem to be measuring this -- the gain from checking these
> extra annotations, with the more precise and/or more timely checks that
> they imply, relative to the erased case where only the language baseline
> contracts are checked. Indeed that's the point of 'exception'
> experiments. Since the biggest effect sizes on display are these ones --
> between Erasure and anything else -- this seems at best distracting. I'd
> be glad to hear from the authors if I'm misunderstanding anything here.
> In the detailed comments below, I have noted some places in the text
> where it would be useful to remind the reader that this design property
> of Typed Racket is at play.

We do think there is a misunderstanding here about what gradual typing is
in general.

1) Erasure is the established name of the semantics implemented by
TypeScript, Flow etc. These languages come with an equally sophisticated
type system as Typed Racket. Programmers add annotations to their program
and the type checker does its best to type check the program in order to
discover mistakes and help the IDE. Then types are erased and the program
is run in the underlying dynamic language. That language's primitive
operations have runtime checks that may detect an incompatible argument
and raise an exception.

@@

2) The other two semantics introduce extra run time checks (in the form of
proxies for Natural and inlined checks for Transient) to enforce that
values untyped code provides to typed code have the expected type. when
one of these checks fails they also issue blame (with different
mechanisms).

3) In Typed Racket programmers cannot add contracts. The precision of the
checks is determined by the types and matches those. In turn the types of
primitive operations match the checks that their Racket counterparts
perform. Finally our benchmarks come with the most precise types that
allow them to type check (where precision is based on subtyping).

@@

> line 380: from this I inferred that "migratory typing" means "gradual
> typing applied at modulewise granularity". Assuming that's correct, it's
> worth saying directly.

There are two points that distinguish gradual typing from migratory
typing: the existence of Dyn and the granularity of components.  We will
clarify the prose.

> line 424: "fully typed correct programs" -- presumably your method could
> also work with not-yet-fully-typed correct programs, just not ranging
> over the entire lattice in those cases. I was wondering whether that
> might give different/interesting results.

Yes, the approach could be applied.
ll: TODO worth a sentence saying that this would probably not be worthwhile to do?
